---
title: "Conditional Statements"
subtitle: ""
description: "Materials adapted from Adrien Osakwe, Larisa M. Soto and Xiaoqi Xie."
---

------------------------------------------------------------------------

## 1. The Logic of Decision Making

In programming, **Conditional Statements** allow your code to branch. Instead of running every line of code from top to bottom, R will only run specific sections **if** a condition is met.

### 1.1 Binary Conditions (`if` ... `else`)

This is used for simple "Yes/No" or "Either/Or" logic.

**The Structure (Pseudo-code):**

```{r}
#| eval: false

if (condition) {
  # Perform this task if condition is TRUE
} else {
  # Perform this task if condition is FALSE
}
```

**Example:**

```{r}
x <- 5

if (x > 5) {
  # If the number is greater than 5, get its square
  x^2
} else {
  # If not, multiply it by 2
  x * 2
}
```

::: callout-note
Because the condition `x > 5` evaluated to **`FALSE`** (since 5 is equal to 5, not strictly greater than 5), R skipped the first block and jumped straight to the `else` block.
:::

### 1.2 Multi-branch Conditions (`else if`)

When your logic is more complex than a simple binary choice, you can chain conditions together. You can have as many `else if` blocks as you need.

**The Structure (Pseudo-code):**

```{r}
#| eval: false

if (condition_1) {
  # Perform Task A
} else if (condition_2) {
  # Perform Task B
} else if (condition_3) {
  # Perform Task C
} else {
  # Perform Task D if NONE of the above were TRUE
}
```

```{r}
x <- 7

if (x > 5 && x < 10) {
  # If x is between 5 AND 10
  x^2
} else if (x < 0 || x > 10) {
  # If x is negative OR greater than 10
  -x
} else {
  # For everything else (like 0, 1, 2, 3, 4, 5)
  x * 2
}
```

## 2. Pro-Tips for Conditionals

### 2.1 `&` vs `&&` (and `|` vs `||`)

You might have noticed the use of double symbols (`&&` and `||`) in the example above. In R, there is a small but important difference:

-   **Single (`&`, `|`):** These are **Vectorized**. Use these when comparing entire columns of data (e.g., filtering a Data Frame).

-   **Double (`&&`, `||`):** These are **Short-circuiting**. They only look at the **first element** of a vector. They are faster and preferred inside `if` statements where you are only checking one value at a time.

### 2.2 The "Missing" Else

The `else` block is optional. If you don't provide one and the `if` condition is `FALSE`, R will simply skip the block and move to the next line of code.

### 2.3 Indentation Matters for Humans!

While R is "space-insensitive" (it doesn't care if your code is a messy pile of text), humans are not. We use **indentation** to show the hierarchy of logic.

It is standard practice to indent the code inside the curly braces `{ }`. This makes it immediately obvious which code belongs to which condition.

::: callout-tip
**The RStudio Secret**

Don't worry about manually hitting the spacebar! RStudio is designed to help you.

-   **Auto-indent:** When you type a `{` and hit Enter, RStudio automatically adds the correct gap.

-   **The "Re-format" Trick:** If your code looks messy, you don't even need to copy-paste. Simply highlight the code and press **`Ctrl + Shift + A`** (Windows) or **`Cmd + Shift + A`** (Mac). RStudio will instantly clean up the spacing and indentation for you.
:::

::: callout-tip
**The `styler` Package**

In professional programming, we often use a package called **`styler`**.

It acts like a "grammar checker" for your code, ensuring it follows the tidyverse style guide used by data scientists at NASA, Google, and beyond.
:::

### 2.4 **Best Practices for Conditional Logic**

#### **The MECE Principle**

When designing your logic, try to follow the **MECE** principle: **Mutually Exclusive, Collectively Exhaustive.**

-   **Mutually Exclusive:** Your conditions shouldn't overlap. A number shouldn't be able to trigger both the `if` and the `else if` block at the same time.

-   **Collectively Exhaustive:** You have considered **all** possibilities. This is why a final `else` block is so important—it acts as a "safety net" for any data you didn't expect.

#### **The Power of Comments**

In bioinformatics, your "Conditions" are often based on scientific thresholds (e.g., p-value \<0.05 or Log2FoldChange \>2).

**Always** write a comment explaining *why* you chose that threshold. Future you will thank you when you're writing the "Methods" section of your paper!
