---
title: "Advanced data manipulation"
subtitle: ""
description: "Materials adapted from Adrien Osakwe, Larisa M. Soto and Xiaoqi Xie."
---

------------------------------------------------------------------------

## 1. The `tidyverse` Philosophy

`dplyr` is part of the **tidyverse**, a collection of R packages designed for data science. The goal is to provide a consistent way to handle data frames.

### The Power of the Pipe (`%>%`)

The pipe operator is the secret sauce of `dplyr`. It takes the output of one function and "pipes" it directly into the next.

-   **Think of it as the word "THEN".**

-   Instead of nesting functions like `head(select(df, year))`, you write: `df %>% select(year) %>% head()`

```{r}
#| message: false
if (!require("gapminder", quietly = TRUE))
    install.packages("gapminder")
library(gapminder)

if (!require("dplyr", quietly = TRUE))
    install.packages("dplyr")
library(dplyr)
```

------------------------------------------------------------------------

## 2. The Core `dplyr` Verbs

### 2.1 `select()`: Picking Columns

Use `select()` when you only want specific variables (columns) from your dataset.

```{r}         
#| collapse: true
# Select specific columns
gapminder %>% 
  dplyr::select(year, country, gdpPercap) %>% 
  head()

# Remove a column using the minus sign
gapminder %>% 
  dplyr::select(-continent) %>% 
  head()
```

### 2.2 `filter()`: Picking Rows

Use `filter()` to find observations (rows) that meet a certain condition.

```{r}         
#| collapse: true
# Include only European countries in the year 2007
gapminder %>% 
  dplyr::filter(continent == "Europe", year == 2007) %>% 
  dplyr::select(country, lifeExp)
```

------------------------------------------------------------------------

## 3. The Power Duo: `group_by()` and `summarize()`

This is the most common workflow in bioinformatics. You **split** the data into groups, **apply** a calculation, and **combine**the results into a summary table.

### 3.1 `group_by()`

This doesn't change the data visually; it creates "hidden" groups that R remembers for the next step.

```{r}         
#| collapse: true
gapminder %>% dplyr::group_by(continent)
```

### 3.2 `summarize()`

This "collapses" each group into a single row of statistics.

```{r}         
#| collapse: true
# Calculate mean GDP and Standard Error for each continent
gapminder %>% 
  dplyr::group_by(continent) %>% 
  dplyr::summarize(
    mean_le = mean(lifeExp),
    se_le = sd(lifeExp) / sqrt(dplyr::n()) # n() counts the number of observations
  )
```

------------------------------------------------------------------------

## 4. `mutate()`: Creating New Variables

`mutate()` allows you to add new columns while keeping the old ones. It is perfect for calculations like unit conversions or normalization.

```{r}         
#| collapse: true
# Create a new column for GDP in billions
gapminder %>% 
  dplyr::mutate(gdp_billion = gdpPercap * pop / 10^9) %>% 
  head()
```

------------------------------------------------------------------------

## 5. Putting It All Together

The true beauty of `dplyr` is chaining everything into one clean pipeline.

```{r}         
#| collapse: true
# A complete pipeline: Create a variable, group it, and summarize it
summary_table <- gapminder %>% 
  dplyr::mutate(gdp_billion = gdpPercap * pop / 10^9) %>% 
  dplyr::group_by(continent, year) %>% 
  dplyr::summarize(
    mean_gdpPercap = mean(gdpPercap),
    mean_pop = mean(pop),
    mean_gdp_billion = mean(gdp_billion)
  )

head(summary_table)
```

### **Pro-Tip: Why use `dplyr::`?**

You’ll notice I used `dplyr::select()` instead of just `select()`. As we discussed in the "Functions" module, many packages have a `filter` or `select` function. Being explicit ensures that your code **never breaks**, even if you load other libraries later.
