---
title: "Hands-on Exercise: Basic Data Manipulation"
subtitle: ""
description: "Materials adapted from Adrien Osakwe, Larisa M. Soto and Xiaoqi Xie."
---

---


```{r}
library(gapminder)
```

## Writing data

Write a data processing snippet to include only the data points collected after 1995 in Asian countries as a CSV file

```{r,eval=F}

```

Separate the `gapminder` data frame into 5 individual data frames, one for each continent. Store those 5 data frames as an `RData` file in the `objects` folder called `continents.RData`.

```{r,eval=FALSE}

```

## Exploring data frames

Finish exploring the `gapminder` data frame and:

-   Find the number of rows and the number of columns\
-   Print the data type of each column\
-   Explain the meaning of everything that `str(gapminder)` prints

```{r,eval=F}

```

In which years has the GDP of Canada been larger than the average of all data points?

```{r,eval=F}

```

Find the mean life expectancy of Switzerland before and after 2000

```{r,eval=F}

```

You discovered that all the entries from 2007 are actually from 2008. Create a copy of the full `gapminder` data frame in an object called `gp`. Then change the year column to correct the entries from 2007.

```{r,eval=F}

```

**Bonus -** Find the mean life expectancy and mean gdp per continent using the function `tapply`

```{r,eval=F}

```

```{r,eval=F}

```
