---
title: "Solutions: Medical Data"
subtitle: ""
description: "Materials adapted from Adrien Osakwe, Larisa M. Soto and Xiaoqi Xie."
---

------------------------------------------------------------------------

```{r}
#| message: false
if (!require("ggplot2", quietly = TRUE))
    install.packages("ggplot2")
library(ggplot2)

if (!require("medicaldata", quietly = TRUE))
    install.packages("medicaldata")
library(medicaldata)

if (!require("dplyr", quietly = TRUE))
    install.packages("dplyr")
library(dplyr)

if (!require("tidyr", quietly = TRUE))
    install.packages("tidyr")
library(tidyr)
```

Use the `covid_testing` data set from package `medicaldata` and everything you've learned so far to answer the following questions:

## Clinics included

**How many clinics participated in the study, and how many valid tests were performed on each one? Did the testing trend vary over time?**

```{r}
covid<-covid_testing
```

### Number of clinics

```{r}
clinics<- covid %>%
          dplyr::select(subject_id,clinic_name,result,pan_day) %>%
          dplyr::distinct()
length(unique(clinics$clinic_name))
```

### Number of valid tests

```{r}
clinics %>%
  dplyr::filter(result!="invalid") %>%
  dplyr::group_by(clinic_name) %>%
  dplyr::summarize(n_test = length(clinic_name)) %>%
  dplyr::arrange(desc(n_test))
```

::: callout-note
**Step-by-Step Breakdown**

1. **`filter(result != "invalid")`**

You are "cleaning the pipe." Before doing any math, you remove the "noise."

-   **The Question:** "Is this a real test result?"

-   **The Action:** Only rows where the result is **NOT** "invalid" are allowed to move to the next step.

2. **`group_by(clinic_name)`**

You are "sorting the mail." R doesn't calculate anything yet; it just creates invisible folders for each clinic.

-   **The Action:** R puts all "Clinic A" rows in one bucket and "Clinic B" rows in another.

3. `summarize(n_test = length(clinic_name))`

You are "collapsing the buckets." This is where the many rows of your original data become one single summary row per group.

-   **The Action:** Inside each clinic's bucket, R counts how many items are there.

-   **Note:** You can also use the shorthand `n()` instead of `length(clinic_name)`—it does the exact same thing!

4. **`arrange(desc(n_test))`**

You are "ranking the results."

-   **The Action:** By default, R sorts low-to-high. `desc()` (descending) flips it so the clinic with the **most** tests appears at the very top.
:::

### Testing trend over time

```{r}
covid %>%
  filter(pan_day<=100) %>%
  group_by(pan_day) %>%
  summarize(n=length(result)) %>%
  ggplot(.,aes(x=pan_day,y=n))+
    geom_point()+
    geom_line()+
    ylab("Number of tests per day")+
    xlab("Pandemic day")
```

## Number of positive tests

**How many patients tested positive vs negative in the first 100 days of the pandemic? Do you notice any difference with the age of the patients? Hint: You can make two age groups and calculate the percentage each age group in positive vs negative tests.**

### Number of positive tests in the first 100 days

```{r}
covid %>%
  filter(result!="invalid" & pan_day<=100) %>%
  group_by(result) %>%
  summarize(n=length(subject_id))
```

### Tests by age group

```{r}
#| message: false
tsts_age<-covid %>%
            filter(result!="invalid" & pan_day<=100) %>%
            mutate(age_group=ifelse(age<=21,"children","adults")) %>%
            group_by(age_group,result) %>%
            summarize(n=length(subject_id)) %>%
            mutate(percent_total=n/sum(n)*100)
tsts_age
```

::: callout-note
**1. The Cleaning Phase (`filter`)**

```{r}
#| eval: false       
filter(result != "invalid" & pan_day <= 100)
```

Before any analysis, you are narrowing the focus to high-quality, early-pandemic data.

-   **`result != "invalid"`**: Removes tests that didn't work (QC failure).

-   **`&`**: A logical "AND." Both conditions must be true.

-   **`pan_day <= 100`**: Only includes data from the first 100 days of the pandemic.

**2. The Classification Phase (`mutate`)**

```{r}
#| eval: false
mutate(age_group = ifelse(age <= 21, "children", "adults"))
```

You are creating a new column "on the fly" to categorize your patients.

-   **`ifelse(test, yes, no)`**: This is a logical switch. If a patient is 21 or younger, they are labeled "children"; otherwise, they are "adults."

**3. The Organization Phase (`group_by`)**

```{r}
#| eval: false
group_by(age_group, result)
```

This is the most critical step. You are telling R to create sub-buckets for every combination of age and test outcome (e.g., "adults-positive," "adults-negative," "children-positive," etc.).

**4. The Aggregation Phase (`summarize`)**

```{r}
#| eval: false
summarize(n = length(subject_id))
```

R collapses the buckets created in the previous step.

-   It counts how many `subject_id` entries are in each bucket and stores that number in a new column called **`n`**.

**5. The Calculation Phase (`mutate`)**

```{r}
#| eval: false
mutate(percent_total = n / sum(n) * 100)
```

This is a "Smart Calculation." Because the data is still grouped by `age_group`, the `sum(n)` happens **separately for each group**.

-   **Example**: For the "adults" rows, it divides their count by the *total number of adults*.

-   This gives you the percentage of results *within* that specific age group, rather than the percentage of the entire dataset.

------------------------------------------------------------------------

**Notes on `sum(n)`**

In this specific code, the `percent_total` will represent the percentage of results **per age group** (e.g., what % of children were positive vs. negative). If you wanted the percentage of the **entire table**, you would need to `ungroup()` the data before this last `mutate`.
:::

## Processing times

**Look at the specimen processing time to receipt, did the sample processing times improve over the first 100 days of the pandemic? Plot the median processing times of each day over the course of the pandemic and then compare the summary statistics of the first 50 vs the last 50 days**

```{r}
covid %>%
  group_by(pan_day) %>%
  dplyr::summarise(median_col_rec_tat=median(col_rec_tat)) %>%
  ggplot(.,aes(x=pan_day,y=median_col_rec_tat)) +
  geom_point()+
  geom_line()
```

```{r}
covid %>%
  mutate(pan_day_group=ifelse(pan_day<50,"first_50","last_50")) %>%
  group_by(pan_day_group) %>%
  dplyr::summarise(mean_col_rec_tat=mean(col_rec_tat),
                   median_col_rec_tat=median(col_rec_tat),
                   min_col_rec_tat=min(col_rec_tat),
                   max_col_rec_tat=max(col_rec_tat))
```

## Bonus - Viral load

**Higher viral loads are detected in less PCR cycles.**

**What can you observe about the viral load of positive vs negative samples.**

**Do you notice anything differences in viral load across ages in the positive samples?**

**Hint: Also split the data into two age groups and try using geom_boxplot()**

```{r,warning=FALSE}
ggplot(covid,aes(y=ct_result,x=result,color=result))+
  geom_boxplot()
```

```{r,warning=FALSE}
covid %>%
  mutate(age_group=ifelse(age<=21,"children","adults")) %>%
  ggplot(.,aes(y=ct_result,x=result,color=age_group))+
    geom_boxplot()
```
