---
title: "Solutions: Advance Data Manipulation"
subtitle: ""
description: "Materials adapted from Adrien Osakwe, Larisa M. Soto and Xiaoqi Xie."
---

------------------------------------------------------------------------

```{r}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(dplyr))
```

Write one command ( can span multiple lines) using pipes that will output a data frame that has only the columns `lifeExp`, `country` and `year` for the records before the year 2000 from African countries, but not for other Continents.

```{r}
tidy_africa <- gapminder %>%
                dplyr::filter(continent == "Africa") %>%
                dplyr::select(year, country, lifeExp)
head(tidy_africa)
```

Calculate the average life expectancy per country. Which country has the longest average life expectancy and which one the shortest average life expectancy?

```{r}
gapminder %>%
   dplyr::group_by(country) %>%
   dplyr::summarize(mean_lifeExp = mean(lifeExp)) %>%
   dplyr::filter(mean_lifeExp == min(mean_lifeExp) | mean_lifeExp == max(mean_lifeExp))
```

In the previous hands-on you discovered that all the entries from 2007 are actually from 2008. Write a command to edit the data accordingly using pipes. In the same command filter only the entries from 2008 to verify the change.

```{r}
gapminder %>%
  dplyr::mutate(year = ifelse(year==2007,2008,year)) %>%
  dplyr::filter(year==2008) %>%
  head()
```
